How the Practice of TDD Influences \\Class Design in Object-Oriented Systems: \\Patterns of Unit Tests Feedback

### 2012-12-22 sbes2012-english/paper.tex
revisao do ricardo

	Despite the fact that Test-Driven Development (TDD) appears to be a software testing practice, 
	many developers affirm that the practice influence on class design. This study aimed
	to better understand the effects of TDD and how the practice influences developers
	during class design on object-oriented systems. 
	We conducted an essentially qualitative exploratory study in which participants
	were invited to implement some exercises using TDD and, based on the data gathered,
	we raised details of how the practice influenced design decisions from the
	participants using interviews.
	We noticed that the practice of TDD drives
	developers during class design by means of constant feedback about its quality. This study
	also named and catalogued feedback patterns perceived by the developers.

@@

The adoption of TDD by the industry has been growing. In a 2010 survey to discover
which practices were used by agile teams, Scott Ambler showed that 53\% of all teams
that replied the survey adopted TDD as a way to validate the work done
\cite{wambler-survey-agile}. Similar numbers can be found on the annual surveys from
Version One that, in the 2012 version \cite{versionone-2012}, showed that
40\% of the respondent teams have been using the practice.

When using TDD, a developer only writes code that is covered by a test. Due
to that, it is common to relate the practice of TDD to software testing. However,
a common discourse among TDD practitioners is its effects on the internal code quality.
Many authors, well-known by industry and academia, such as Kent Beck \cite{TDDByExample}, 
Robert Martin \cite{agile-ppp}, Steve Freeman \cite{GOOS}, and Dave Astels \cite{astels-tdd}, 
state (without scientific evidence) that the practice of TDD promotes a significative
improvement on class design, helping developers to create more cohesive less coupled
classes.

On the other hand, the way in which the practice of TDD guides the developer
through class design is not clear. We observed that in our qualitative study
with TDD practitioners, in a Brazilian agile event. We interviewed
ten participants about TDD effects \cite{aniche-wbma} and, for our surprise,
none of them could satisfactory explain how the practice guides them to better
designs.
Siniaalto and Abrahamsson \cite{alarming-results} also share this opinion. They also
noticed that the effects of TDD can not be as automatic as expected by most people.
The related work, discussed in Section \ref{cap:trabalhos-relacionados},
only evaluates if the practice of TDD makes difference in the produced code. Only a few
of them contain a qualitative study that details how the practice does that difference.
With this information, developers would know how to leverage the practice of TDD in order to 
better create class design.

@@

Based on the fact that the software development process involves many
human factors and is totally sensitive to the context that it belongs,
we conducted an essentially qualitative exploratory study in which professionals
from industry were invited to implement some prepared exercises using TDD and,
based on the gathered data, we investigated the details of how the practice influenced
the participants' class design decisions by means of interviews.

@@

TDD is a software development practice that is based on a repetition of a small
cycle. First, developers write a failing test. Then, they make it pass,
by implementing the desired functionality. After that, they refactor the code
to remove code or data duplication that was generated by the process. Besides that,
simplicity is intrinsic to the process: the practitioner writes the simplest
test that fails and writes the simplest implementation that makes the
test pass. This cycle is also known as "Red-Green-Refactor", as it remembers the
colors that a developer usually sees when doing TDD: the red means that a test
is failing, and green means that the test passes successfully.

@@

Authors such as Kent Beck \cite{aim-fire}, Dave Astels \cite{astels-tdd}, and
Robert Martin \cite{bob-martin} state that TDD is, in fact, a class design
technique \cite{tdd-taxonomy} \cite{aim-fire}.
In their opinion, the change in the traditional development cycle, although very simple,
aggregates many other benefits to the produced code: more simplicity, less coupling, and
a high cohesion. Ward Cunningham, one of the Extreme Programming pioneers, summarizes
the discussion in one statement: \textit{"Test-First programming is not a testing technique"} 
\cite{aim-fire}.
According to Janzen, a more clear definition is the one that says that TDD is the 
art of producing automated tests to production code, and use this process to guide
the class design and programming \cite{agilealliance-tdd} \cite{tdd-taxonomy}.

It is a known fact that class design tend to lose quality during the software evolution \cite{evolution-lehman}.
However, it is hard to discuss quality in terms of class design. In this study context,
we used the SOLID principles, detailed in Martin's work \cite{bob-martin}, which are
mentioned in the following section.

@@

Many empirical studies have been done to evaluate the effects of TDD.
In most of them, the effects of the practice in the class design is not taken
into account, and only the effect on external quality is measured.
In contrary to what this study proposes, many of them chosed
to have a bigger control in the experiment. They were also conduted of the academic
environment, with students that belong to the most different computer courses.

@@

The study done by George e Williams \cite{george-e-williams} showed that,
although TDD can initially reduce the productivity of the more inexperienced
developers, a qualitative analysis showed that 92\% of the developers think
that TDD helps to maintain the quality of the code, and 79\% believes that
the practice promotes a simpler class design.

@@

Li \cite{angela-li} proposed a qualitative study to understand the efficacy of TDD.
In a case study, she collected the perceptions of TDD practitioners about the
benefits of the practices. To achieve that, she interviewed five developers
from software companies in Auckland, Australia. The results of the interviews
were analyzed and discussed in terms of code quality, software quality, and
programmer's productivity. Regarding code quality, Li concluded that TDD
guides developers to simpler and better designed classes. In addition, the main
factors that contribute to these benefits are the confidence to refactor and
modify code, a higher code coverage, a deeper understanding of the requirements,
the ease of understanding code, and the elevated
satisfaction of the developers.

TDD practitioners usually makes uses of other agile practices, such as
pair programming. This makes the evaluation of the practice more difficult.
Madeyski \cite{madeyski-package-dependencies} observed the results
among groups that practiced TDD, groups that practiced pair programming,
and the combination between them, and he was not able to show a significant
difference between teams that use TDD and teams that use pair programming
in terms of class package dependency management. However, when combining
the results, he found that TDD can help managing dependencies in class level.
According to him, developers should use TDD, but they should be aware of 
problems in class design.

Muller and Hagner \cite{muller-e-hagner} showed that TDD does not result in
better quality or productivity. Steinberg \cite{steinberg} showed that
code produced with TDD is more cohesive and less couple. Students also reported
that defect were easier to fix.

@@

As presented, only a few studies evaluated the effects of TDD on class design.
And, when they did, they only discussed what the effects are and not exactly
\textbf{how} TDD has influenced them. Josefsson \cite{josefsson}, in his discussion
about the need of an architectural phase and the effects of TDD in this matter,
comes to the same conclusion. According to him, studies about TDD that are
found in the current literature are very limited and, because of that, the
"known" effects of the practice in class design can not be explained. Based on
our literature review, we believe that this limitation still exists.

Most part of these studies also does not take into account the experience
of the developer who is practicing TDD. Usually they only discuss it
during threats to validity. Janzen \cite{janzen-phd}, in his PhD, perceived
that mature developers obtain more benefits from the practice, writing simpler
classes. Besides, mature developers tend to choose TDD more often than
less experienced developers.

However, studies that analyze TDD from the class design point of view does not
come to a conclusive results; many of them even affirm that the effects of TDD
is not that different from those teams that do not practice it. Even the Janzen's PhD
thesis was inconclusive regarding the influence of the practice in coupling
and cohesion \cite{janzen-phd}.

@@

To achieve our goal, we chose to conduct an essentially 
exploratory qualitative study with developers in industry. Participants
were invited to implement a set of pre-prepared exercises. After that,
they were interviewed about how the practice influenced them in their
class design decisions. This section details the planning as well as
the data analysis process.

@@

The main goal of this study was \textbf{to understand the relationship between
the practice of TDD and class design decisions taken by programmers in object-oriented
systems}. To achieve it, we answered the following questions:

@@

Participants from different software development companies in the Brazilian
industry were selected. The participants' profile is discussed in sub-section
\ref{sec:planejamento-participantes}. All of them were invited to implement
a few problems using Java in a defined timeframe. Participants practiced
TDD in one problem and did not practice in the other. The problems,
as well as the instruction of using or not TDD, were
randomized in order to reduce bias.

@@

To calculate these metrics, we implemented our own tool. The reason is that
most part of the existent tools make use of compiled code (instead of pure source
code). When analyzing many projects at once, it is not an easy task to compile all
of them as each one has its own way of being build.
Our tool contains an automated test suite and is released as open source
\footnote{\url{http://www.metricminer.org.br}. 
Last access on July, the 6th, 2012.}.

@@

At the end of the exercise, all participants filled out a survey about their performance
on the exercises. After that, an initial analysis filtered the participants that
were interviewed posteriorly. The selection of the participants were based on
their answers on the survey plus the produced source code. Participants that implemented
a good solution on one problem but not on the other were selected. Besides, participants
that mentioned possible effects of the practice on class design were also interviewed.

@@

Since the decisions taken by a programmer during the class design activity can be
influenced by many different factors, the questions were made to make participants
to triangulate their answers and to make them isolate the practice of TDD from
the other possible factors of influence. Participants that did not articulate well
the answers were removed from the analysis process.

@@

Developers from the Brazilian software development industry were invited to be part
of this research. The participants were evaluated according to their experience in TDD,
software development, Java, and unit tests. The only requirement was that the participant
should already knew how to write a unit test.

@@

The numbers are a little different when talking about their experience in software
development. 24\% of them worked with software development for the last 4 or 5 years.
28\% of them worked from 6 to 10 years. Only 20\% of them developed software for less
than 2 years. 64\% of the participants use Java. However, 36\% of them do not use
Java in their daily work. All of them know JUnit and 64\% use mock objects 
\footnote{Mock objects are objects created during a unit test. They mock the behavior 
of a concrete object. Usually they are used to isolate the unit test from the other classes. 
More information about it can be found on \cite{mocks}.}
during their development activities.
Only 12\% has never heard about mock objects. When talking about object-orientation,
in the open question, most part of the participants affirmed to have a good experience
on it. Some of them even affirmed to master the subject. Only a few of them affirmed
to have basic knowledge about OO.

@@

The statistical test used was the Wilcoxon. It is a non parametric hypothesis test
used to compare two samples and to check whether there is a difference in the mean
of both samples. Therefore, we used Wilcoxon to compare the mean between
the code metrics in code produced with and without TDD. The significance level
used was the standard (0.05).

@@

In Table \ref{metricas-industria}, we show the \textit{p-values} found for
the difference between codes produced with and without TDD. By the numbers,
we observe that there was no significant difference in cyclomatic complexity
and efferent coupling.
Lack of cohesion of methods was different in two exercises (1 and 4). A difference
also appeared on the quantity of lines in a method (exercise 4) and quantity of
methods (exercise 1).

When looking to the data of all exercises together, no metric presented a significant
difference. We concluded that, at least quantitatively, the practice of TDD did not
make any difference in code metrics.

@@

In Table \ref{valores-separados}, we calculated the \textit{p-values} of the metric,
splliting them by experience in software development and TDD. The values for the group that
was experienced in TDD and not experienced in software development was not calculated as no
participant fits on it.

By the numbers, we perceived that cohesion was the only metric that presented a significant
different for developers that are experienced in software development and TDD.

@@

The values presented before corroborate with many of the related work.
Apparently TDD does not influence the code to the point of changing
the metric values for coupling, cohesion, and simplicity. However, this is
incoherent to the industry common feeling of the positive effects of TDD on class design.
As stated, in this study we conducted a qualitative analysis to understand
how this influence goes from the point of view of the developers.

@@

An interesting point is that the participants, regardless their experience on TDD or
software development, commented similar points. Because of that, we did not separate
discussions in the categories presented in Section \ref{sec:planejamento}.

@@

Different from what was expected, a majority of the participants affirmed that the
practice of TDD would not change the class design produced by them during the exercises
they did not use the practice.
The main justification was that experience and previous knowledge on object-orientation
guided them during the class design. All participants agreed that a developer with no
knowledge in any of the mentioned areas would create a good class design only by practicing
TDD.

Two good examples that reinforce the point were given by the participants. One of them
said that made use of a design pattern \cite{gof} that he has learned a few days before. Another
participant mentioned that his studies on SOLID principles helped him during the exercises.
The transcription below were extracted from the interviews:

@@

In addition, the only participant from industry that has never practiced TDD before
affirmed that he did not feel any improvement in the class design when practicing.
Curiously, this participant said that he already knew that "TDD was a design technique".
It somehow indicates that the popularity of the effects of TDD in class design
is big.
That opinion was slightly different from the experienced participants, 
who affirm that TDD is not only about design, but also about testing.

However, in contrary to the fact that TDD does not guide developers directly to a
good class design, all participants believe that TDD has positive effects on class design.
Many of them even mentioned the difficulty when trying to stop using TDD:

@@

The majority of the participants also commented that one difference that is
perceived when they practice TDD is the constant feedback. In traditional testing,
the time between the production code writing and test code writing is too big.
When practicing TDD, developers are forced to write the test first, and receive all
feedback a test can give sooner.

@@

Reduce the time between the code writing and test writing also helps developers to
create code that effectively solves the problem. According to the participants, in traditional
testing, developers write too much code before actually knowing if it works:

\begin{framed}
	\textit{"[The test] is not only a specification; it should actually work. So, as you really reduce
	the time between writing a software that works and to test it, you end up perceiving if that
	part works or not quicker (...)"}
\end{framed}

@@

That point was already raised by Feathers \cite{feathers-synergy}.
The harder is to write the test, the bigger the chance of a class design problem.
According to him, there is a big synergy between a high testable class and
a good class design: if developers is looking for testability, they end up
creating a good class design; if they are looking for a good class design, they end
up writing testable code.

In that search for testability, developers are encouraged to write code
that is easily testable. Codes like that contain a few interesting
characteristics, such as the ease of invoke the expected behavior,
the non need of complicated pre-conditions, and the explicit declaration
of all dependencies.

The participants went deep further. During the interviews, many of them
mentioned about many patterns they have found in unit tests. These patterns
make them think about possible design problems in the class they are building.
These patterns are discussed in the following sub-sections.

@@

As mentioned before, most part of the feedback a test gives occurs in the moment
that the developer finds a difficulty when writing the test. This sub-section discusses
patterns mentioned by the participants that make them believe that there is a
problem in the class design.

@@

One pattern that was not explicitly mentioned by the participants, but
noticed by us, is when a developer feels the need to test a method
that is not public. Private methods usually helps to transform a public method
into something more easier to be read. When trying to test it in an isolated
way, the developer may have found a method that contains too much responsibility.
That responsibility can be moved to a different class. We call it \textbf{Tests in Non Public Methods}.

@@

The abusive use of mock objects to test a single class indicates that
the class under test has coupling issues. It is possible to infer that a class
that makes use of many mock objects depends upon many classes and, therefore,
tend to be an unstable class. We call it \textbf{Too Many Mock Objects}.

@@

The lack of an abstraction usually makes that a simple change needs to
be done in different pieces of code. When requirements change and
developers are obligated to do the same change in different tests,
that indicate a lack of a correct abstraction to represent the problem
and avoid the code repetition. We call it \textbf{Same Change in Different Tests}.
Analogously, developers can perceive the same thing when they create the
same tests for different entities. We call it \textbf{Repeated Tests for Different Entities}.

@@

%% ------------------------------------------------------------------------- %%
\subsection{Relationship between Patterns and Class Design Principles}

@@

The proposed exercises were small compared to a real project. However, all exercises
contain localized class design problems. Once that this study evaluates the effects
of TDD on class design, we believe the exercises can simulate real design problems
satisfactory.
Besides, at the end of the implementation, participants replied a question about the
similarity between the exercises and real world problems. All of them affirmed that
the problems found in the exercises were very closed to the ones they deal in real world.

@@

%% ------------------------------------------------------------------------- %%
\subsubsection{Reserarcher's influence}

Researchers are a fundamental key in qualitative research. They can influence
the interpretation of the results because of past experiences.
In this study, our opinion had a strong influence on the participants selection
to the interview.
Besides, the exercises were created by us and, somehow, it may have influenced
the TDD practiced. To reduce this problem, we reviewed all analysis, looking for
non clear or incorrect conclusions.

@@

In this study context, a possible bias would be a participant that replies
what the literature says, and not exactly what s/he feels about the practice.
To reduce the bias, we would eliminate participants that replied questions superficially.
In practice that did not happen. Only a few answers were superficially (and these
ones were eliminated).

@@

Although we contacted many different companies and software development groups,
the number of participants invited may not be enough to generalize the findings.

@@

In this study, we discussed and understood how the TDD practice can make the
difference in classes design.
In addition, besides endorsing the quantitative studies form the literature, this study
goes beyond and analyzes feedback patterns that appeared in the moment when
the developer uses TDD and, in practice, guides him during the design.

@@

A possible future work is the developing of tools that automatically detects this patterns
and warn the developer about a possible code smell in design. Besides that,
the search for more patterns like the ones shown here can improve even more the feedback
provided by the practice.

@@

The practice of TDD \textbf{can} affect the creating process of class design. However,
unlike the comments made by the industry, \textbf{the TDD practice cannot guide the
developer to a good class design automatically}; the experience and knowledge of
developer are essential to develop of oriented object software.

@@

\textbf{Therefore, this is the way how the practice guides the developer to
a better class design; giving constant feedback about potential issues in
class design. It is a developer's task to realize this issues and, 
improve the design. }

@@

When writing a unit test to a specific class, the developer is forced to follow
the same steps: writing of the scenario, execution of the action being
tested, and assuring that the behavior was executed like the
expected.
\textbf{Any trouble writing this steps, can imply issues in the class design}. 
The aware developer, notices and improves the respectively design class.

@@

We thank to the companies that participated in this study: Caelum, Bluesoft,
Amil e WebGoal (São Paulo and Poços de Caldas). Besides, we thank the
developers that took part of this research independently.

>>>>

	Despite the fact that Test-Driven Development (TDD) appears to be a software testing practice, 
	many developers affirm that the practice influences on class design. This study aimed
	to better understand the effects of TDD and how the practice influences developers
	during class design on object-oriented systems. 
	We conducted an essentially qualitative exploratory study in which participants
	were invited to implement some exercises using TDD and, based on the data gathered,
	we raised details of how the practice influenced design decisions from the
	participants using interviews.
	We noticed that the practice of TDD drives
	developers during class design by means of constant feedback about its quality. This study
	also named and catalogued feedback patterns perceived by the developers.

@@

The adoption of TDD by the industry has been growing. In a 2010 survey to discover
which practices were used by agile teams, Scott Ambler showed that 53\% of all teams
that replied to the survey adopted TDD as a way to validate the work done
\cite{wambler-survey-agile}. Similar numbers can be found on the annual surveys from
Version One that, in the 2012 version \cite{versionone-2012}, showed that
40\% of the respondent teams have been using the practice.

When using TDD, a developer only writes code that is covered by a test. Due
to that, it is common to relate the practice of TDD to software testing. However,
a common discourse among TDD practitioners is regarding its effects on the internal code quality.
Many authors, well-known by industry and academia, such as Kent Beck \cite{TDDByExample}, 
Robert Martin \cite{agile-ppp}, Steve Freeman \cite{GOOS}, and Dave Astels \cite{astels-tdd}, 
state (without scientific evidence) that the practice of TDD promotes a significative
improvement on class design, helping developers to create more cohesive less coupled
classes.

On the other hand, the way in which the practice of TDD guides the developer
through class design is not clear. This was observed in our qualitative study
with TDD practitioners, in a Brazilian agile event. We interviewed
ten participants about TDD effects \cite{aniche-wbma} and, to our surprise,
none of them could satisfactorily explain how the practice guides them to better
designs.
Siniaalto and Abrahamsson \cite{alarming-results} also share this opinion. They also
noticed that the effects of TDD can not be as automatic as expected by most people.
The related work, discussed in Section \ref{cap:trabalhos-relacionados},
only evaluates if the practice of TDD makes any difference in the produced code. Only a few
of them contain a qualitative study that details how the practice makes that difference.
With this information, developers would know how to leverage the practice of TDD in order to 
better create class design.

@@

Based on the fact that the software development process involves many
human factors and is totally sensitive to the context that it belongs to,
we conducted an essentially qualitative exploratory study in which professionals
from the industry were invited to implement some prepared exercises using TDD and,
based on the gathered data, we investigated the details of how the practice influenced
the participants' class design decisions by means of interviews.

@@

TDD is a software development practice that is based on a repetition of a small
cycle. First, developers write a failing test. Then, they make it pass,
by implementing the desired functionality. After that, they refactor the code
to remove code or data duplication that was generated by the process. Besides that,
simplicity is intrinsic to the process: the practitioner writes the simplest
test that fails and writes the simplest implementation that makes the
test pass. This cycle is also known as "Red-Green-Refactor", as it recalls the
colors that a developer usually sees when doing TDD: the red means that a test
is failing, and green means that the test passes successfully.

@@

Authors such as Kent Beck \cite{aim-fire}, Dave Astels \cite{astels-tdd}, and
Robert Martin \cite{bob-martin} state that TDD is, in fact, a class design
technique \cite{tdd-taxonomy} \cite{aim-fire}.
In their opinion, the change in the traditional development cycle, although very simple,
aggregates many other benefits to the produced code: more simplicity, less coupling, and
a high cohesion. Ward Cunningham, one of the Extreme Programming pioneers, summarizes
the discussion in one statement: \textit{"Test-First programming is not a testing technique"} 
\cite{aim-fire}.
According to Janzen, a clearer definition is the one that says that TDD is the 
art of producing automated tests to production code, and using this process to guide
the class design and programming \cite{agilealliance-tdd} \cite{tdd-taxonomy}.

It is a known fact that class design tends to lose quality during the software evolution \cite{evolution-lehman}.
However, it is hard to discuss quality in terms of class design. In this study context,
we used the SOLID principles, detailed in Martin's work \cite{bob-martin}, which are
mentioned in the following section.

@@

Many empirical studies have been conducted to evaluate the effects of TDD.
In most of them, the effects of the practice in the class design are not taken
into account, and only the effect on external quality is measured.
Contrary to what this study proposes, many of them chose
to have a bigger control in the experiment. They were also conduted of the academic
environment, with students that belong to different computer courses.

@@

The study made by George and Williams \cite{george-e-williams} showed that,
although TDD can initially reduce the productivity of the more inexperienced
developers, a qualitative analysis showed that 92\% of the developers think
that TDD helps to maintain the quality of the code, and 79\% believe that
the practice promotes a simpler class design.

@@

Li \cite{angela-li} proposed a qualitative study to understand the efficacy of TDD.
In a case study, she collected the perceptions of TDD practitioners about the
benefits of the practices. To achieve that, she interviewed five developers
from software companies in Auckland, New Zealand. The results of the interviews
were analyzed and discussed in terms of code quality, software quality, and
programmer's productivity. Regarding code quality, Li concluded that TDD
guides developers to simpler and better designed classes. In addition, the main
factors that contribute to these benefits are the confidence to refactor and
modify code, a higher code coverage, a deeper understanding of the requirements,
the ease of understanding code, and the elevated
satisfaction of the developers.

TDD practitioners usually make use of other agile practices, such as
pair programming. This makes the evaluation of the practice more difficult.
Madeyski \cite{madeyski-package-dependencies} observed the results
among groups that practiced TDD, groups that practiced pair programming,
and the combination between them, and he was not able to show a significant
difference between teams that use TDD and teams that use pair programming
in terms of class package dependency management. However, when combining
the results, he found that TDD can help managing dependencies at class level.
According to him, developers should use TDD, but they should be aware of 
problems in class design.

Muller and Hagner \cite{muller-e-hagner} showed that TDD does not result in
better quality or productivity. Steinberg \cite{steinberg} showed that
code produced with TDD is more cohesive and less coupled. Students also reported
that defects were easier to fix.

@@

As presented, only a few studies evaluated the effects of TDD on class design.
And, when they did, they only discussed what the effects are and not exactly
\textbf{how} TDD has influenced them. Josefsson \cite{josefsson}, in his discussion
about the need for an architectural phase and the effects of TDD in this matter,
comes to the same conclusion. According to him, studies about TDD that are
found in the current literature are very limited and, because of that, the
"known" effects of the practice in class design can not be explained. Based on
our literature review, we believe that this limitation still exists.

Most of these studies also do not take into account the experience
of the developer who is practicing TDD. Usually they only discuss it
during threats to validity. Janzen \cite{janzen-phd}, in his PhD, perceived
that mature developers obtain more benefits from the practice, writing simpler
classes. Besides, mature developers tend to choose TDD more often than
less experienced developers.

However, studies that analyze TDD from the class design point of view does not
come to conclusive results; many of them even affirm that the effects of TDD
are not so different from those teams that do not practice it. Even the Janzen's PhD
thesis was inconclusive regarding the influence of the practice in coupling
and cohesion \cite{janzen-phd}.

@@

To achieve our goal, we chose to conduct an essentially 
exploratory qualitative study with developers in the industry. Participants
were invited to implement a set of pre-prepared exercises. After that,
they were interviewed about how the practice influenced them in their
class design decisions. This section details the planning as well as
the data analysis process.

@@

The main goal of this study was \textbf{to understand the relation between
the practice of TDD and class design decisions taken by programmers in object-oriented
systems}. To achieve it, we answered the following questions:

@@

Participants from different software development companies in the Brazilian
industry were selected. The participants' profiles are discussed in sub-section
\ref{sec:planejamento-participantes}. All of them were invited to implement
a few problems using Java in a defined timeframe. Participants practiced
TDD in one problem and did not practice in the other. The problems,
as well as the instruction of using or not TDD, were
randomized in order to reduce bias.

@@

To calculate these metrics, we implemented our own tool. The reason is that
most of the existent tools make use of compiled code (instead of pure source
code). When analyzing many projects at once, it is not an easy task to compile all
of them as each one has its own way of being built.
Our tool contains an automated test suite and is released as open source
\footnote{\url{http://www.metricminer.org.br}. 
Last access on July, the 6th, 2012.}.

@@

At the end of the exercise, all participants filled out a survey about their performance
on the exercises. After that, an initial analysis filtered the participants that
were interviewed posteriorly. The selection of the participants was based on
their answers on the survey plus the produced source code. Participants that implemented
a good solution on one problem but not on the other were selected. Besides, participants
that mentioned possible effects of the practice on class design were also interviewed.

@@

Since the decisions taken by a programmer during the class design activity can be
influenced by many different factors, the questions were made to make participants
triangulate their answers and to make them isolate the practice of TDD from
the other possible factors of influence. Participants that did not articulate well
the answers were removed from the analysis process.

@@

Developers from the Brazilian software development industry were invited to be part
of this research. The participants were evaluated according to their experience in TDD,
software development, Java, and unit tests. The only requirement was that the participant
should already know how to write a unit test.

@@

The numbers were a little different when talking about their experience in software
development. 24\% of them worked with software development for the last 4 or 5 years.
28\% of them worked from 6 to 10 years. Only 20\% of them developed software for less
than 2 years. 64\% of the participants use Java. However, 36\% of them do not use
Java in their daily work. All of them know JUnit and 64\% use mock objects 
\footnote{Mock objects are objects created during a unit test. They mock the behavior 
of a concrete object. Usually they are used to isolate the unit test from other classes. 
More information about it can be found on \cite{mocks}.}
during their development activities.
Only 12\% have never heard about mock objects. When talking about object-orientation,
in the open question, most of the participants affirmed to have a good experience
on it. Some of them even affirmed to master the subject. Only a few of them affirmed
to have basic knowledge about OO.

@@

The statistical test used was the Wilcoxon. It is a non parametric hypothesis test
used to compare two samples and to check whether there is a difference in the meaning
of both samples. Therefore, we used Wilcoxon to compare the meaning between
the code metrics in code produced with and without TDD. The significance level
used was the standard (0.05).

@@

In Table \ref{metricas-industria}, we show the \textit{p-values} found for
the differences between codes produced with and without TDD. By the numbers,
we observe that there was no significant differences in cyclomatic complexity
and efferent coupling.
Lack of cohesion of methods was different in two exercises (1 and 4). A difference
also appeared on the quantity of lines in a method (exercise 4) and quantity of
methods (exercise 1).

When looking at the data of all exercises together, no metric presented a significant
difference. We concluded that, at least quantitatively, the practice of TDD did not
make any difference in code metrics.

@@

In Table \ref{valores-separados}, we calculated the \textit{p-values} of the metric,
splliting them by experience in software development and TDD. The values for the group that
was experienced in TDD and not experienced in software development was not calculated as no
participant fitted on it.

By the numbers, we perceived that cohesion was the only metric that presented a significant
difference for developers that are experienced in software development and TDD.

@@

The values presented above corroborate with many of the related work.
Apparently TDD does not influence the code to the point of changing
the metric values for coupling, cohesion, and simplicity. However, this is
incoherent to the industry's common feeling of the positive effects of TDD on class design.
As stated, in this study we conducted a qualitative analysis to understand
how this influence goes from the point of view of the developers.

@@

An interesting point is that the participants, regardless of their experience on TDD or
software development, commented on similar points. Because of that, we did not separate
discussions in the categories presented in Section \ref{sec:planejamento}.

@@

Different from what was expected, the majority of the participants affirmed that the
practice of TDD would not change the class design produced by them during the exercises
in which they did not use the practice.
The main justification was that experience and previous knowledge on object-orientation
guided them during the class design. All participants agreed that a developer with no
knowledge in any of the mentioned areas would create a good class design only by practicing
TDD.

Two good examples that reinforce the point were given by the participants. One of them
said that he made use of a design pattern \cite{gof} that he had learned a few days before. Another
participant mentioned that his studies on SOLID principles helped him during the exercises.
The transcription below was extracted from the interviews:

@@

In addition, the only participant from the industry that had never practiced TDD before
affirmed that he did not feel any improvement in the class design when practicing.
Curiously, this participant said that he already knew that "TDD was a design technique".
It somehow indicates that the popularity of the effects of TDD in class design
is big.
That opinion was slightly different from the experienced participants, 
who affirmed that TDD was not only about design, but also about testing.

However, contrary to the fact that TDD does not guide developers directly to a
good class design, all participants believe that TDD has positive effects on class design.
Many of them even mentioned the difficulty when trying to stop using TDD:

@@

The majority of the participants also commented that one difference that is
perceived when they practice TDD is the constant feedback. In traditional testing,
the time between the production code writing and test code writing is too long.
When practicing TDD, developers are forced to write the test first, and receive all
feedback a test can give sooner.

@@

Reducing the time between the code writing and test writing also helps developers to
create a code that effectively solves the problem. According to the participants, in traditional
testing, developers write too much code before actually knowing if it works:

\begin{framed}
	\textit{"[The test] is not only a specification; it should actually work. So, as you really reduce
	the time between writing a software that works and to test it, you end up by perceiving if that
	part works or not quicker (...)"}
\end{framed}

@@

That point had already been raised by Feathers \cite{feathers-synergy}.
The harder it is to write the test, the bigger the chance of a class design problem.
According to him, there is a big synergy between a high testable class and
a good class design: if developers are looking for testability, they end up by
creating a good class design; if they are looking for a good class design, they end
up by writing testable code.

In that search for testability, developers are encouraged to write code
that is easily testable. Codes like that contain a few interesting
characteristics, such as the ease of invoking the expected behavior,
the non need of complicated pre-conditions, and the explicit declaration
of all dependencies.

The participants went even further. During the interviews, many of them
mentioned about many patterns they have found in unit tests. These patterns
make them think about possible design problems in the class they are building.
These patterns are discussed in the following sub-sections.

@@

As mentioned before, most of the feedback a test gives occurs at the moment
when the developer finds a difficulty when writing the test. This sub-section discusses
patterns mentioned by the participants that make them believe that there is a
problem in the class design.

@@

One pattern that was not explicitly mentioned by the participants, but
noticed by us, is when a developer feels the need to test a method
that is not public. Private methods usually help to transform a public method
into something easier to be read. When trying to test it in an isolated
way, the developer may have found a method that contains too much responsibility.
That responsibility can be moved to a different class. We call it \textbf{Tests in Non Public Methods}.

@@

The abusive use of mock objects to test a single class indicates that
the class under test has coupling issues. It is possible to infer that a class
that makes use of many mock objects depends upon many classes and, therefore,
tends to be an unstable class. We call it \textbf{Too Many Mock Objects}.

@@

The lack of an abstraction usually makes that a simple change needs to
be done in different pieces of code. When requirements change and
developers are obligated to do the same change in different tests,
that indicates a lack of a correct abstraction to represent the problem
and avoid the code repetition. We call it \textbf{Same Change in Different Tests}.
Analogously, developers can perceive the same thing when they create the
same tests for different entities. We call it \textbf{Repeated Tests for Different Entities}.

@@

%% ------------------------------------------------------------------------- %%
\subsection{Relation between Patterns and Class Design Principles}

@@

The proposed exercises were small compared to a real project. However, all exercises
contain localized class design problems. Once this study evaluates the effects
of TDD on class design, we believe the exercises can simulate real design problems
satisfactorily.
Besides, at the end of the implementation, participants answered a question about the
similarity between the exercises and real world problems. All of them affirmed that
the problems found in the exercises were very similar to the ones they deal with in real world.

@@

%% ------------------------------------------------------------------------- %%
\subsubsection{Researcher's influence}

Researchers are a fundamental key in qualitative research. They can influence
the interpretation of the results because of past experiences.
In this study, our opinion had a strong influence on the participants selection
to the interview.
Besides, the exercises were created by us and, somehow, this may have influenced
the TDD practiced. To reduce this problem, we reviewed all analysis, looking for
non clear or incorrect conclusions.

@@

In this study context, a possible bias would be a participant that replies
what the literature says, and not exactly what s/he feels about the practice.
To reduce the bias, we would eliminate participants that replied questions superficial.
In practice that did not happen. Only a few answers were superficially (and these
ones were eliminated).

@@

Although we contacted many different companies and software development groups,
the number of participants invited may not have been enough to generalize the findings.

@@

In this study, we discussed and understood how the TDD practice can make the
difference in classes design.
In addition, besides endorsing the quantitative studies from the literature, this study
goes beyond and analyzes feedback patterns that appeared in the moment when
the developer uses TDD and, in practice, guides him during the design.

@@

A possible future work is the developing of tools that automatically detect these patterns
and warn the developer about a possible code smell in design. Besides that,
the search for more patterns like the ones shown here can improve even more the feedback
provided by the practice.

@@

The practice of TDD \textbf{can} affect the creating process of class design. However,
unlike the comments made by the industry, \textbf{the TDD practice cannot guide the
developer to a good class design automatically}; the experience and knowledge of
the developer are essential to develop oriented object software.

@@

\textbf{Therefore, this is the way how the practice guides the developer to
a better class design; giving constant feedback about potential issues in
class design. It is a developer's task to realize these issues and, 
improve the design. }

@@

When writing a unit test to a specific class, the developer is forced to follow
the same steps: writing of the scenario, execution of the action being
tested, and assuring that the behavior was executed as
expected.
\textbf{Any trouble writing these steps can imply issues in the class design}. 
The attentive developer notices and improves the respective design class.

@@

We thank the companies that participated in this study: Caelum, Bluesoft,
Amil e WebGoal (São Paulo and Poços de Caldas). Besides, we thank the
developers that took part in this research independently.
